Evaluation of Cache-Oblivious Data Structures

### 2008-05-10 paper/paper.tex
Added some initial test configurations.

% Marielle suggested:
%  - give the "standard" memory model an explicit name
%  - move example of cache oblivious data structures to front
%  - mention examples of cache unaware and cache aware algorithms

% Note explicitly that cache-oblivious algorithms are relatively complex, so
% that the main question is wether the improvedlocality of reference 
% compensates for the overhead incurred.

@@

State space search is used to verify the correctness of software programs. For
this purpose, programs are first modeled using a formal language, that is also
used to specify properties of the program that should hold during its
execution. A program can be in a (possibly infinite) amount of states, one of
which is usually designated the \emph{initial state}. If a transition from one
state to another is possible (according to the rules of the formal language
used) the latter state is said to be a \emph{successor state} of the former.
Generating the successors of a particular state is also called
\emph{expanding} the state. Of course, the model must be non-deterministic to
allow more than one successor to exist for a single state.

@@

% TODO: note that the algorithm uses the queue, successors and set, and that
the set is the bottleneck.

% TODO: different name?

% Describe case: state search, set data structure.
% Describe (globally) what to measure: execution time

@@

% - NIPS VM
% - SPIN model checker
% - Promela-to-NIPS compiler

@@

@@

\section{Metrics}
% Describe how performance was measured
% Describe data set used

>>>>

% Marielle suggested:
%  - give the "standard" memory model an explicit name
%    (isn't "cache-unaware" enough?)
%  - move examples of cache oblivious data structures to front
%  - mention examples of cache unaware and cache aware algorithms
%    (at least Btree, tries and hash tables are mentioned)

% Note explicitly that cache-oblivious algorithms are relatively complex
% resulting in higher constant consts, so that the main question is wether
% the improved locality of reference compensates for the overhead incurred.

@@

State space search is used to verify the correctness of software programs. For
this purpose, programs are first modeled using a formal language, that is also
used to specify properties of the program that should hold during its
execution. An executing program can be in a (possibly infinite) amount of states,
one of which is usually designated the \emph{initial state}. If a transition
from one state to another is possible (according to the rules of the formal
language used) the latter state is said to be a \emph{successor state} of the
former.
Generating the successors of a particular state is also called
\emph{expanding} the state. Of course, the model must include some form of
non-determinism to allow more than one successor to exist for a single state.
In practice, this non-determinism usually comes from processes that execute in
parallel, where the precise interleaving of the (observed) execution of
instructions in these processes is non-deterministic, unless synchronization
primitives (such as channels, semaphores, atomic execution blocks, et cetera)
are used to enforce a particular ordering.

@@

Depending on how states are inserted and extracted from the queue, the search
may be breadth-first (meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps) or depth-first (meaning that states that are generated most recently
by the successor function are expanded first). In the first case, the queue
operates by a first-in first-out principle, while in the second case the queue
works in a last-in first-out (stack) principle. In general, breadth-first
search requires a much larger queue capacity than depth-first search.

From the pseudo-code it is clear that the state space search algorithm does not
accomplish much by itself; instead, the real work is done by the successor
function and the queue and the set datastructures. Queues can easily be
implemented efficiently (adding and removing states takes O(1) time).
The efficiency of the successor function depends on the execution model used,
but is typically linear in the number of states produced. In practice,
therefore, the bottleneck of the algorithm is the set data structure.

@@

% Describe (globally) what to measure: execution time
% - NIPS VM
% - SPIN model checker
% - Promela-to-NIPS compiler

@@

The NIPS VM represents program state as a binary string; the size of the state
depends on the number of active processes in the program, and is typically in
the order of a few hundred bytes.

@@

\section{Metrics}
% Describe how performance was measured
%  (mock set implementation; times/sizes subtracted)

% Describe data set used; per model:
%   average state size
%   total size
%   ratio of insertions/queries

% Note that state size depends on number of processes

The following models are used for benchmarking:
\begin{itemize}
\item\verb#eratosthenes# is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers upto a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only by exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
also varies during execution.

\item\verb#leader2# is a leader election algorithm with a configurable number
of processes competing for leadership ($N$). Since a constant number of
processes will be created and these processes cannot make progress until all
of them are created, almost all states have the same size.

\item\verb#peterson_N# is a Peterson's solution to the multi-process mutual
exclusion problem (using global variables instead of channels for
communication) with a configurable number of processes ($N$). The processes
are created atomically, so the state size is constant after initialization.

% mobile1
% pftp
% snoopy
% sort

% TODO: table of models with parameters,

\end{itemize}
